%% Cover Letter for Nature Genetics
%% Path-Probability Models Outperform Point-Estimate Scores for Noncoding GWAS Gene Prioritization

\documentclass[11pt]{letter}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}

\signature{Abduxoliq Ashuraliyev\\Independent Researcher\\Tashkent, Uzbekistan\\Jack00040008@outlook.com}

\address{Abduxoliq Ashuraliyev\\Independent Researcher\\Tashkent, Uzbekistan}

\begin{document}

\begin{letter}{Editorial Office\\Nature Genetics}

\opening{Dear Editors,}

I am submitting the manuscript ``Path-Probability Models Outperform Point-Estimate Scores
for Noncoding GWAS Gene Prioritization'' for consideration at \textit{Nature Genetics}.

\textbf{What we did:}
We introduce mechanism graphs---a probabilistic inference object that explicitly represents
causal chains from GWAS variants through regulatory elements to genes, tissues, and traits,
with calibrated uncertainty at each step.

\textbf{Why this is a field-level advance:}
\begin{itemize}
\item \textbf{New inference object:} Mechanism graphs are not a pipeline but a formal
      probabilistic representation with explicit generative assumptions (noisy-OR aggregation)
      that maintains probability semantics throughout---``0.8 means 80\% true'' holds at
      each module, not just final output.
\item \textbf{Multi-causal colocalization:} Integration of coloc.susie addresses the
      single-causal-variant assumption that causes standard colocalization to fail
      at 34\% of multi-signal loci.
\item \textbf{Functional enhancer--gene links:} Ensemble of Activity-by-Contact and
      PCHi-C provides ground-truth-validated regulatory connections beyond distance,
      improving prioritization at complex loci where distance alone is uninformative.
\end{itemize}

\textbf{What is validated independently:}
\begin{itemize}
\item \textbf{Anti-leak benchmarks:} Three-tier evaluation with explicit provenance
      tracking---Tier 1 Mendelian genes verified absent from all L2G training sets;
      76\% recall at rank 20 vs 58\% for Open Targets L2G.
\item \textbf{Enhancer--gene links:} Validated against 847 CRISPRi-derived regulatory
      connections (AUPRC 0.71 vs 0.54 for distance-only).
\item \textbf{Per-module calibration:} Expected Calibration Error $< 0.05$ at each stage,
      demonstrated with reliability curves and bootstrap confidence intervals.
\item \textbf{Cross-study replication:} 78\% of GTEx colocalization signals replicate
      in independent eQTL Catalogue datasets ($r = 0.89$ effect size correlation).
\end{itemize}

\textbf{What is released:}
\begin{itemize}
\item GitHub: \url{https://github.com/ProgrmerJack/Mechanism-GWAS-Causal-Graphs}
\item Zenodo DOI: \href{https://doi.org/10.5281/zenodo.17798899}{10.5281/zenodo.17798899}
      (published, open access, MIT license)
\item Mechanism atlas for 2,134 cardiometabolic loci
\item Deterministic Snakemake pipeline with versioned dependencies
\item Data manifests with SHA256 checksums for all inputs
\end{itemize}

\textbf{Why Nature Genetics readers will care:}
\begin{itemize}
\item \textbf{Target discovery:} Interpretable mechanism paths prioritize specific
      regulatory hypotheses for experimental follow-up, not just gene rankings.
\item \textbf{Functional genomics:} Calibrated probabilities enable principled
      resource allocation---high-confidence paths justify costly perturbation experiments.
\item \textbf{Statistical genetics:} Uncertainty propagation and cross-study replication
      address systematic concerns about single-study eQTL dependence and benchmark leakage.
\end{itemize}

\textbf{Preprint disclosure:}
The Zenodo archive (DOI: 10.5281/zenodo.17798899) serves as the preprint record.
This work has not been submitted for peer review elsewhere. Upon publication,
the preprint record will be updated with the journal DOI per Nature policy.

\textbf{Suggested reviewers:}
\begin{enumerate}
\item \textbf{Chris Wallace} (University of Cambridge) --- coloc developer; expertise
      in QTL colocalization and multi-causal fine-mapping
\item \textbf{Jesse Engreitz} (Stanford University) --- ABC Model developer; expertise
      in enhancer--gene linking and regulatory genomics
\item \textbf{Hilary Finucane} (Broad Institute) --- PoPS developer; expertise in
      gene prioritization and polygenic methods
\item \textbf{Alexander Gusev} (Dana-Farber Cancer Institute) --- TWAS/colocalization
      expertise; known for rigorous statistical evaluation
\item \textbf{Kaur Alasoo} (University of Tartu) --- eQTL Catalogue lead; expertise
      in cross-study QTL replication
\item \textbf{Tuuli Lappalainen} (New York Genome Center) --- expertise in QTL
      interpretation and functional follow-up; provides rigorous methodological critique
\end{enumerate}

\textbf{Competing interests:} None declared.

\textbf{Data and code availability:} All code and data are publicly available
without restrictions as detailed in the manuscript.

\closing{Sincerely,}

\end{letter}
\end{document}
